Synthesis of Linear Functional Programs

### 2021-07-14 article/article.tex
tweaks

With this section we aim to formalize the implicit \todo{mais uma vez\dots} rules that guide
the synthesis of more \todo{nao se faz sintese de specs, mas sim a
  partir de specs} expressive specifications, alongside the already
explicit rules that model the core of the synth \todo{synthesis} process, resulting in
a \emph{sound} set of rules that hopes to serve as a source of ideas
\todo{ nao me parece\dots As regras caracterizam o que esta
  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots }
on taming synthesis of recursive functions with richer types in a
linear context, as well as describe the system in enough detail for
the synthetiser to be reproducible by a theory-driven implementation.

@@

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synth a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synth a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x \rightarrow M$ and the output linear context is
$\Delta'$.

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either
instance \todo{instantiate}
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs \todo{ Na escrita, repetition
  does not legitimize. Constructors can take as arguments values of
  the type they are defining, ou algo assim}. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
\lolli C$, and its derivation:
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of Bar}\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of Bar}\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of Bar}\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    \te{Foo} \lolli C \Uparrow}
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $\te{Foo}$ is also infinite.
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
    \Downarrow}
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. \todo{Não se percebe muito bem a
  segunda parte da frase, rephrase!}
For this, we need two more contexts, $\Rho_C$
for restrictions \todo{constraints} on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
These modifications prevent the derivation from not ending
\todo{infinite derivations} for the scenarios described above,
however, they also greatly limit the space of derivable programs,
leaving the synthetiser effectively unable to synth specifications
with recursive types\todo{refrasear tendo em conta que se sintetiza a
  partir de specs e nao specs}, which is so undesirable that we add three rules
to complement the restrictions on construction and destruction of
recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

\todo[inline]{esta frase nao me parece muito necessaria}
The ADT rules draw closer to a final form, but will still need a revision
after the introduction of polymorfism and polymorfic parametrization on ADTs.

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synth a function of type $A \lolli B$ named
\emph{foo}, the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

Having said that, we won't allow the use of the recursive call in $\Gamma$ right
away. If we did, every synth goal would have a trivial proof (a non-terminating
function that just calls itself) that could even shadow other valid, relevant,
solutions.  Instead, our framework only allows the use of recursion after having
deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
hypothesis as arguments. To illustrate, having $\textrm{data List} =
\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
"available" when deconstructing ($\textrm {List} \vdash C$), and only in
the \emph{Cons} branch. % (TODO: rewrite)

% TODO: E como evitamos coisas trivialmente estúpidas?

\todo[inline]{Polymorphic}

\mypara{Polymorphic Types} A polymorphic specification is a type $X$ \todo{T} and a set of
bound variables that quantify universally the type variables in $X$, called a
\emph{scheme} ($\forall \overline{\alpha}.\ \tau$).
\todo[inline]{Não é um tipo e um conjunto de variáveis bound. É um
  tipo da forma $\forall \overline{\alpha}.\ T$ em que $\dots$ }
Synthesis of a scheme is straightforward. \todo{Diria que nao e
  especialmente straightforward.}
First, \textbf{type variables are considered atomic types}, then, instantiate
the bound vars\todo{variables} of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
\lolli \alpha$ could be instanced \todo{instantiated} to
$\alpha0 \lolli \alpha0$. We add a rule for this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
a type variable can instance a goal that's a type variable of the same name),
corresponds to the synthesis of a program where some expressions are treated
agnostically -- nothing constrains their type, meaning the program is
polymorphic * TODO: rewrite this?.\todo[inline]{Use an example.}

>>>>

With this section we aim to formalize the rules that guide
synthesis from more expressive specifications, alongside the already
well defined rules that model the core of the synthesis process, resulting in
a \emph{sound} set of rules that hopes to serve as a source of ideas
\todo{ nao me parece\dots As regras caracterizam o que esta
  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots}
  \todo{R: não parece ser sound ?}
on taming synthesis of recursive functions with richer types in a
linear context, as well as describe the system in enough detail for
the synthetiser to be reproducible by a theory-driven implementation.

@@

Putting together linear logic, the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synth a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synth a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x \rightarrow M$ and the output linear context is
$\Delta'$.

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT $T$ and its constructors stand for any ADT with form
\[
    \textrm{data}\ T = C_1\ X_1\ \vert\ C_2\ X_2\ \vert\
\dots\ \vert\ C_n\ X_n
\]

% repetition does not legitimize :p

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs \todo{ Na escrita, repetition does not
legitimize. Constructors can take as arguments values of the type they are
defining, ou algo assim}. Seemingly small, this change has a big impact in the
synthesis process. Take, for instance, the ADT defined as $\textrm{data}\ T =
C_1\ T$, the synth goal $T \lolli C$, and its derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Symmetrically, the
derivation for the synth goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, simetrically, don't allow construction of an
ADT when trying to synth an argument for its constructor. \todo{R: A segunda parte confusa era da frase de
  cima?, está melhor?}
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    T\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
%
These modifications prevent the infinite derivations in the scenarios described
above, however, they also greatly limit the space of derivable programs, leaving
the synthetiser effectively unable to synth from specifications with recursive
types\todo{refrasear tendo em conta que se sintetiza a partir de specs e nao
specs R: refrasear mas só com "from" não fica bem?}, which is so undesirable that we add three rules to complement the
restrictions on construction and destruction of recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synth a function of type $A \lolli B$ named
\emph{f}, the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($f{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

Having said that, we won't allow the use of the recursive call in $\Gamma$ right
away. If we did, every synth goal would have a trivial proof (a non-terminating
function that just calls itself) that could even shadow other valid, relevant,
solutions.  Instead, our framework allows the use of recursion only after having
deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
hypothesis as arguments. To illustrate, having $\textrm{data List} =
\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
"available" when deconstructing ($\textrm {List} \vdash C$), and only in
the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)

\todo[inline]{Polymorphic R: ??}

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
bound variables that quantify universally the type variables in $\tau$, which is
also called a \emph{scheme}.
Synthesis of a scheme comprises of turning it into a non-quantified type, and
then treating its type variables agnostically. \todo{R: Melhor que
straightforward?}
First, \textbf{type variables are considered atomic types}, then, instantiate
the bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
\lolli \alpha$ could be instantiated to
$\alpha0 \lolli \alpha0$. We add a rule for this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
%
As such, the construction of a derivation in which, to some types, no specific
rules can be applied (except for the \textsc{init} rule, a type variable can
instance a goal that's a type variable of the same name), corresponds to the
synthesis of a program where some expressions are treated agnostically --
nothing constrains their type -- meaning the program is polymorphic. The
simplest example is the polymorphic function \emph{id} of type $\forall \alpha
.\ \alpha \lolli \alpha$. The
program synthed from that specification is $\lambda x \rightarrow x$, a lambda
abstraction that doesn't constrain its parameter $x$ in any way. \todo{R :
rewrite this? não estranho?}
